---
title: "external"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{external}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

Using external storrs.

Often it is useful to retrieve data from an external resource
(especially websites).  The way this works is:

1. We do a key lookup on the storr; if that succeeds (i.e. it maps
to a hash) continue as normal..

2. If the lookup fails, pass the key (and namespace) to a "hook"
function that generates an R object (in any way).

This is in some ways a variant on the memoisation pattern; if the
key refers to a set of arguments to a long running function we get
something like memoisation (see the bottom of this file).

As an example, this vignette will download some DESCRIPTION files
from github, using the name of the repository as the key.

The first step is writing a hook function; this is a function with
arguments `(key, namespace)` that returns an R object.  For
packages stored in the root directory of a repository we can build
URLs of the form

```
https://raw.githubusercontent.com/<username>/<repo>/master/DESCRIPTION
```

So if the key is a username/repo pair and we ignore namespace we
can write a function:
``` {r }
fetch_hook_gh_description <- function(key, namespace) {
  if (!isTRUE(unname(capabilities("libcurl")))) {
    stop("This vignette requires libcurl support in R to run")
  }
  fmt <- "https://raw.githubusercontent.com/%s/master/DESCRIPTION"
  path <- tempfile("gh_description_")
  on.exit(file.remove(path))
  code <- download.file(sprintf(fmt, key), path, mode = "wb")
  if (code != 0L) {
    stop("Error downloading file")
  }
  as.list(read.dcf(path)[1, ])
}
```

This function downloads the requested DESCRIPTION file into a
temporary file (which it promises to delete later using `on.exit`),
checks that the download was successful, then reads in the
downloaded file and converts it into a list.

The `httr` and `curl` packages make this a little easier to do with
authorisation so that this would work for private repositories by
using a token.

With this in place, we can build a storr:
``` {r }
st <- storr::storr_external(storr::driver_environment(),
                            fetch_hook_gh_description)
```

The first argument here is a storr *driver* (i.e., a `driver_`
function).  If you have a storr that you want to use, pass it as
`st$driver` to extract the underlying driver (and share storage
with your existing storr).

As with other storr creation functions, you can set the default
namespace using the `default_namespace` argument.

The returned object is exactly the same as a usual storr except
that the `get` method has changed (this is done by inheritence).
The `get` method only behaves differently when the object is not
present in the storr, in which case it will try to fetch the object
and insert it into the storr.

At first there is nothing in here:
``` {r }
st$list()
```

But we can still `get` things from the storr:
``` {r }
d <- st$get("richfitz/storr")
```

Once a key has been fetched, it will be retrieved locally:
``` {r }
identical(st$get("richfitz/storr"), d)
```

And it will be present within the storr, as shown by `list`:
``` {r }
st$list()
```

If an external resource cannot be located, storr will throw an
error of class `KeyErrorExternal`:
``` {r }
tryCatch(st$get("richfitz/no_such_repo"),
         KeyErrorExternal = function(e)
           message(sprintf("** Repository %s not found", e$key)))
```

This would happen for all errors, including lack of internet
connectivity, corrupt file downloads, etc.  The original error will
be returned as the `$e` element of the error if you need to
distinguish between types of failure.  The `KeyErrorExternal` is
also a `KeyError` class, so code that catches `KeyErrors` will
still work as expected.

For more details on storr exception handling, see the `storr`
vignette (`vignette("storr", package = "storr")`)

Note that if you want to persist the storage of the descriptions
you would need to mangle the key:
``` {r }
st_rds <- st$export(storr::storr_rds(tempfile(), mangle_key = TRUE))
st_rds$list()
st_rds$get("richfitz/storr")$Version
```

The `st_rds` storr does not include the fetch hook; it is a plain storr.

``` {r }
st_rds$destroy()
```

# Memoisation

The external storr can support a form of memoisation, though it
might be simpler to implement this directly (see below).

Suppose you have some expensive function `f(a, b)`
``` {r }
f <- function(a, b) {
  message(sprintf("Computing f(%.3f, %.3f)", a, b))
  ## ...expensive computation here...
  list(a, b)
}
```

and a set of parameters to run the function over, with each
parameter set (row) associated with an id:
``` {r }
pars <- data.frame(id = as.character(1:10), a = runif(10), b = runif(10),
                   stringsAsFactors = FALSE)
```

The `hook` here simply looks the parameters up and arranges to run them:
``` {r }
hook <- function(key, namespace) {
  p <- pars[match(key, pars$id), -1]
  f(p$a, p$b)
}

st <- storr::storr_external(storr::driver_environment(), hook)
```

The first time the result is retrieved the message will be printed
(the function is evaluated)
``` {r }
x <- st$get("1")
```

The second time, it will not be as the result is retrieved from the
storr:
``` {r }
identical(st$get("1"), x)
```

This idea can be generalised by storing the *parameters* and the
*functions* in the storr so that we lose the dependency on the
global variables:
``` {r }
st <- storr::storr_environment()
st$set("experiment1", pars, namespace = "parameters")
st$set("experiment1", f, namespace = "functions")

hook2 <- function(key, namespace) {
  f <- st$get(namespace, namespace = "functions")
  pars <- st$get(namespace, namespace = "parameters")
  p <- pars[match(key, pars$id), -1]
  f(p$a, p$b)
}

st_use <- storr::storr_external(st$driver, hook2)

x1 <- st_use$get("1", "experiment1")
x2 <- st_use$get("1", "experiment1")
```

Memoisation in the style of the `memoise` package is possible to
implement, but is not provided in the package.  Implementation is
straightforward and will work with any driver:
``` {r }
memoise <- function(f, driver = storr::driver_environment()) {
  force(f)
  st <- storr::storr(driver)
  function(...) {
    ## NOTE: also digesting the inputs as a key here (in addition to
    ## storr's usual digesting of values)
    key <- digest::digest(list(...))
    tryCatch(
      st$get(key),
      KeyError = function(e) {
        ans <- f(...)
        st$set(key, ans)
        ans
      })
  }
}
```

Here's a function that will print when it is evaluated
``` {r }
f <- function(x) {
  message("computing...")
  x * 2
}
```

Create the memoised function
``` {r }
g <- memoise(f)
```

The first time an argument is seen, `f()` will be run, printing a message
``` {r }
g(1)
```

Subsequent times will be looked up from the storr:
``` {r }
g(1)
```

Storr takes about twice as long as memoise (memoise does a direct
key->value mapping rather than going through hashed values because
it is the only thing that ever touches its cache).  However, the
overhead is approximately half of one call to `message()` so it's
not that bad.
