---
title: "storr"
author: "Rich FitzJohn"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{storr}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

``` {r }
library(storr)
```

At the moment, `storr` requires a driver to be explicitly given;
that will change to be a bit more user-friendly shortly.

The idea with different drivers is that different ways of storing
data have different trade-offs in terms of speed, size,
concurrency, etc; the approach here should allow different
backends to be switched in and out while keeping the same
user-facing interface.

`driver_rds` stores contents at some path by saving out to rds
files.  Here I'm using a temporary directory for the path; the
driver will create a number of subdirectories here.
``` {r }
path <- tempfile("storr_")
dr <- storr::driver_rds(path)
```

With this driver object we can create the `storr` object which is
what we actually interact with:
``` {r }
st <- storr::storr(driver=dr)
```

# Key-value store:

The most simple way of interacting with a `storr` object is
`get`/`set`/`del` for getting, setting and deleting data stored at
some key.  To store data:
``` {r }
st$set("mykey", mtcars)
```

To get the data back
``` {r }
head(st$get("mykey"))
```

What is in the `storr`?
``` {r }
st$list()
```

Or, much faster, test for existance of a particular key:
``` {r }
st$exists("mykey")

st$exists("another_key")
```

To delete it:
``` {r }
st$del("mykey")
```

It's gone!
``` {r }
st$list()
```

# Lists and indexable serialisation

A disadvantage of saving R objects to disk is you have to read the
entire thing in at once.  `storr` addresses one specific solution
to this problem; allowing simple indexing of objects that are
list-like (i.e., things you could throw at `lapply` productively).

Here's a daft object that we want to serialise but address by index
later.
``` {r }
set.seed(1)
obj <- setNames(lapply(1:5, runif), letters[1:5])
obj
```

Rather than use `set`, use `set_list` to assign the object against
a key.
``` {r }
st$set_list("mylist", obj)
```

The object appears in the `storr` as before
``` {r }
st$list()
```

and can be accessed in its entirety:
``` {r }
st$get("mylist")
```

but it has a different type:
``` {r }
st$type("mylist")
```

Can be queried on length:
``` {r }
st$length_list("mylist")
```

To access an individual element:
``` {r }
st$get_list_element("mylist", 2)
```

To access multiple elements (names are dropped at present, but
might be retrieveable as an option in future)
``` {r }
st$get_list_elements("mylist", c(1, 3, 5))
```

To assign to individual elements:
``` {r }
st$set_list_element("mylist", 3, "a different value")
st$get_list_element("mylist", 3)
```

Retrieve the whole list again; names are still there but the value
of the 3rd element has changed.
``` {r }
st$get("mylist")
```

Lists can be deleted as before
``` {r }
st$del("mylist")
```

# Import / export

Objects can be imported in and exported out of a `storr`;

Import from a list, environment or another `storr`
``` {r }
st$import(list(a=1, b=2))
st$list()
st$get("a")
```

Export to an environment or another `storr`
``` {r }
e <- new.env(parent=emptyenv())
st$export(e)
ls(e)
e$a
```

Convenience function that does the same as the above (exports to a
new environment that has `.GlobalEnv` as its parent)
``` {r }
e2 <- st$to_environment()
ls(e2)
parent.env(e2)

st2 <- storr::storr(driver=storr::driver_rds(tempfile("storr_")))
st2$list()
st2$import(st)
st2$list()
```

# Supported backends

* environments (`driver_environment`) - mostly for debugging and
transient storage, but by far the fastest.
* on disk with rds (`driver_rds`) - zero dependencies, quite fast,
will suffer under high concurrency because there is no file
locking.
* Redis (`driver_redis`) - uses
[`redux`](https://github.com/richfitz/redux) to store
the data in a [Redis](http://redis.io) database.  Slower than rds,
but can allow multiple R processes to share the same set of objects.
* rlite (`driver_rlite`) - stores data in an
[rlite](https://github.com/seppo0010/rlite) using
[`rrlite`](https://github.com/ropensci/rrlite).  This is the
slowest at present and does not support concurrency at all.  But
rlite has the potential to be as useful as SQLite is so this will
improve.

# Implementation details

`storr` includes a few useful features that are common to all
drivers.

## Content addressable lookup

The only thing that is stored against a key is the hash of some
object.  Each driver does this a different way, but for the rds
driver it stores small text files that list the hash in them.  So:
``` {r }
dir(file.path(st$driver$path_keys, "objects"))
readLines(file.path(st$driver$path_keys, "objects", "a"))
st$get_hash("a")
```

Then there is one big pool of hash / value pairs:
``` {r }
st$list_hashes()
```

in the rds driver these are stored like so:
``` {r }
dir(file.path(st$driver$path_data))
```

This is going to need garbage collecting every so often - no
reference counting is done so stale objects can build up.
``` {r }
st$gc()
st$list_hashes()
dir(file.path(st$driver$path_data))
```

Eventually this might be automated but for now it's not.

## Environment-based caching

Every time data passes across a `get` or `set` method, `storr`
stores the data in an environment within the `storr` object.
Because we store the content against its hash, it's always in sync
with what is saved to disk.  That means that the look up process
goes like this:

1. Ask for a key, get returned the hash of the content
2. Check in the caching environment for that hash and return that
if present
3. If not present, read content from disk/db/wherever the driver
stores it and save it into the caching environment

Because looking up data in the environment is likely to be orders
of magnitide faster than reading from disks or databases, this
means that commonly accessed data will be accessed at a similar
speed to native R objects, while still immediately reflecting
changes to the content (because that would mean the hash changes)

To demonstrate:
``` {r }
st <- storr::storr(driver=storr::driver_rds(tempfile("storr_")))
```

This is the caching environent; currently empty
``` {r }
ls(st$envir)
```

Set some key to some data:
``` {r }
set.seed(2)
st$set("mykey", runif(100))
```

The environment now includes an object with a *name* that is the
same as the *hash* of its contents:
``` {r }
ls(st$envir)
```

Extract the object from the environment and hash it
``` {r }
storr:::hash_object(st$envir[[ls(st$envir)]])
```

When we look up the value stored against key `mykey`, the first
step is to check the key/hash map; this returns the key above (this
step *does* involve reading from disk)
``` {r }
st$get_hash("mykey")
```

It then calls `$get_value` to extract the value associated with
that hash - the first thing that function does is try to locate the
hash in the environment, otherwise it reads the data from wherever
the driver stores it.
``` {r }
st$get_value
```

The speed up is going to be fairly context dependent, but 10x seems
pretty good in this case (some of the overhead is simply a longer
code path as we call out to the driver).
``` {r }
hash <- st$get_hash("mykey")
microbenchmark::microbenchmark(st$get_value(hash, use_cache=TRUE),
                               st$get_value(hash, use_cache=FALSE))
```
